---
title: "Bellabet Project"
author: "Ogheneovo"
date: "2023-04-21"
output: html_document
---
### Loading libraries
```{r}
library("tidyverse")
library("skimr")
library("janitor")
library("dplyr")
library("ggplot2")
library("ggpubr")
library(ggcorrplot)
```

### Loading data (csv file) into r 
```{r}
daily_activity <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
heartrate <- read_csv("Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")
sleepday <- read_csv("Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weightlog_info <- read_csv("Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
```

### Viewing and Cleaning Data

```{r}
head(daily_activity)
head(heartrate)
head(sleepday)
head(weightlog_info)
```

### Getting the column names
```{r}
colnames(daily_activity)
colnames(heartrate)
colnames(sleepday)
colnames(weightlog_info)
```

### Checking for the number of columns, rows and datatype()
```{r}
glimpse(daily_activity)
glimpse(heartrate)
glimpse(sleepday)
glimpse(weightlog_info)
```

### Check for missing values and duplicates 
```{r}
print ("Count of missing variables - ")
sum(is.na(daily_activity))
print(" Count of Duplicate values - ")
sum(duplicated(daily_activity))
```

```{r}
print ("Count of missing variables - ")
sum(is.na(heartrate))
print(" Count of Duplicate values - ")
sum(duplicated(heartrate))
```

```{r}
print ("Count of missing variables - ")
sum(is.na(sleepday))
print(" Count of Duplicate values - ")
sum(duplicated(sleepday))
```

### Removing duplicate rows from the sleepday dataset
```{r}
sleepday <- sleepday %>% distinct()
print ("Count of missing variables - ")
sum(is.na(sleepday))
print(" Count of Duplicate values - ")
sum(duplicated(sleepday))
```

```{r}
print ("Count of missing variables - ")
sum(is.na(weightlog_info))
print(" Count of Duplicate values - ")
sum(duplicated(weightlog_info))
```
### Check to confirm rows with NA values and decide if we remove them or create a new df
```{r}
colSums(is.na(weightlog_info))
```

***The fat column contains 65 NA values out of 67 entries. We will need to remove the fat column as it's not useful to our analysis.***

### Removing the NA column from our weightlog_info dataset
```{r}
weightlog_info <- weightlog_info[,-5]
colSums(is.na(weightlog_info))
```

## Processing Data

### Seperating columns having data & time for the different dataset so as to get unifomity of date
```{r}
heartrate <- separate(heartrate, Time, into = c('Date', 'Time'), sep = ' ')
heartrate <- heartrate %>% select(-Time)
```

```{r}
sleepday <- separate(sleepday, SleepDay, into = c('Date', 'Time'), sep = ' ')
sleepday <- sleepday %>% select(-Time)
```

```{r}
weightlog_info <- separate(weightlog_info, Date, into = c('Date', 'Time'), sep = ' ')
weightlog_info <- weightlog_info %>% select(-Time)
```

### Renaming the ActivityDate column from daily_activity dataset so as to harmonize the names
```{r}
daily_activity <- daily_activity %>% rename(Date = ActivityDate)
```

### Checking for columns names
```{r}
colnames(daily_activity)
colnames(heartrate)
colnames(sleepday)
colnames(weightlog_info)
```

### Checking the datatype if they are correct

```{r}
str(daily_activity)
str(heartrate)
str(sleepday)
str(weightlog_info)
```

### Changing the Date columns from character to Date

```{r}
daily_activity <- mutate(daily_activity, Date = as.Date(Date, format = "%m/%d/%Y"))
heartrate <- mutate(heartrate, Date = as.Date(Date, format = "%m/%d/%Y"))
sleepday <- mutate(sleepday, Date = as.Date(Date, format = "%m/%d/%Y"))
weightlog_info <- mutate(weightlog_info, Date = as.Date(Date, format = "%m/%d/%Y"))
```

### Changing the Id column from numeric to character
```{r}
daily_activity$Id <- as.character(daily_activity$Id)
heartrate$Id <- as.character(heartrate$Id)
sleepday$Id <- as.character(sleepday$Id)
weightlog_info$Id <- as.character(weightlog_info$Id)
```

### Checking if the change is effected
```{r}
class(daily_activity$Date)
class(heartrate$Date)
class(sleepday$Date)
class(weightlog_info$Date)
```

### Create the average activity covered by each individual within the timeframe
```{r}
average_activity <- daily_activity %>% 
                      group_by(Id) %>% 
                      summarise(Avg_distance = mean(TotalDistance),Avg_steps = mean(TotalSteps),
                      Avg_calories = mean(Calories))
head(average_activity)
unique(average_activity$Id)
```

### Visualize persons that covered the most distance and those with the least distance within the period

```{r}
top_dist_users <- average_activity %>%
       arrange(desc(Avg_distance)) %>% 
       slice(1:10) %>%
       ggplot(., aes(x = Id, y=Avg_distance)) + 
       geom_bar(stat="identity", fill="deepskyblue3") + 
       ggtitle('Top 10 Users with Highest Distance Covered')
top_dist_users + coord_flip()
```

```{r}
least_dist_users <- average_activity %>%
       arrange(Avg_distance) %>% 
       slice(1:10) %>%
       ggplot(., aes(x = Id, y=Avg_distance)) + 
       geom_bar(stat="identity", fill="deepskyblue3") + 
       ggtitle('Top 10 Users with least Distance Covered')
least_dist_users + coord_flip()
```

### Average Calories users generated within the period
```{r}
avg_calories <- average_activity %>%
       arrange(desc(Avg_calories)) %>% 
       slice(1:10) %>%
       ggplot(., aes(x = Id, y=Avg_calories)) + 
       geom_bar(stat="identity", fill="deepskyblue3") + 
       ggtitle('Users with Highest Calories Expended')
avg_calories + coord_flip()
```

*Top IDs burns an average between 2500 and 3500 calories on a daily basis*

### Average steps covered by users within the period
```{r}
avg_steps <- average_activity %>%
       arrange(desc(Avg_steps)) %>% 
       slice(1:10) %>%
       ggplot(., aes(x = Id, y=Avg_steps)) + 
       geom_bar(stat="identity", fill="deepskyblue3") + 
       ggtitle('Users with Most Steps Covered')
avg_steps + coord_flip()
```

***For users with most steps, a minimal average of 9000 steps is taken on a daily basis***

### Understanding daily steps, distance covered and calories burnt within the period
```{r}
daily_activities <- daily_activity %>% 
                      group_by(Date) %>% 
                      summarise(daily_distance = mean(TotalDistance), daily_steps = mean(TotalSteps), 
                      daily_calories = mean(Calories))
head(daily_activities)
```

### Daily distance covered
```{r}
daily_dist_covered <- ggplot(daily_activities, aes(x=Date, y=daily_distance,)) + 
  geom_line(size = 1.4)+
  ggtitle("Daily Distance Covered")+
  xlab("Date")+
  scale_x_date(date_labels="%b %d", date_breaks = "4 day")
  ylim(0,10)
daily_dist_covered
```

### daily steps covered
```{r}
daily_steps_covered <- ggplot(daily_activities, aes(x=Date, y=daily_steps,)) + 
  geom_line(size = 1.4)+
  ggtitle("Daily Steps Covered")+
  xlab("Date")+
  scale_x_date(date_labels="%b %d", date_breaks = "4 day")
  ylim(5000,8500)
daily_steps_covered
```

### Daily calories burnt
```{r}
daily_calories <- ggplot(daily_activities, aes(x=Date, y=daily_calories,)) + 
  geom_line(size = 1.4)+
  ggtitle("Daily Calories Expended")+
  xlab("Date")+
  scale_x_date(date_labels="%b %d", date_breaks = "4 day")+
  ylim(2100,2500)
daily_calories
```


```{r}
steps_calories <- daily_activities %>% 
  ggplot(aes(x = daily_steps, y = daily_calories, color = daily_calories)) +
  geom_point() +
  geom_smooth(method = "lm", color = "gray50", se = FALSE, size = 1) +
  ggtitle("Daily Steps Vs Daily Calories Burnt") +
  xlab("Daily Steps") +
  ylab("Daily Calories") +
  scale_color_gradient(low = "red", high = "green", limits = c(6000, 9000)) +
  guides(color = guide_colorbar(title = "Daily_Calories")) +
  xlim(6000, 9000)+
  ylim(2100,2500)+
  theme_light() +
  theme(legend.title = element_text(size = 12, face = "bold")) + 
  theme(plot.caption = element_text(hjust = 0, margin = margin(t = 10)))
steps_calories
```

***The above charts reveals there exist a relationship between steps taken and calories expended on a daily basis.***

### Understaning the weight and BMI of different IDs and classifying them into categories to know number of persons under each category.
```{r}
weight_info <- weightlog_info %>% 
                      group_by(Id) %>% 
                      summarise(Avg_weight = mean(WeightKg),Avg_BMI = mean(BMI))
head(weight_info)
```


### Adding a new column to determine IDs that are underweight, healthy, overweight and obese
```{r}
weight_info$Weight_status <- with(weight_info, ifelse(Avg_BMI > 30.0, 'obese',
                         ifelse(Avg_BMI > 25, 'overweight', 
                         ifelse(Avg_BMI > 18, 'healthy', 'underweight'))))
head(weight_info)
```

### Average weight Vs BMI to show the relationship
```{r}
Weight_BMI <- ggplot(data = weight_info) + geom_smooth(mapping = aes(x=Avg_weight, y=Avg_BMI)) +
  geom_point(mapping = aes(x=Avg_weight, y=Avg_BMI)) +
  labs(title = "Chart of Average weight Vs BMI", x = "Average weight (Kg)", y = "Average BMI")
Weight_BMI
```


### Visualising to Know different weight status of the IDs
```{r}
weight_info %>% 
  pull(Weight_status) %>% 
  fct_count() %>% 
  rename(Weight_status = f, count = n) %>% 
  ggdonutchart("count", label = "Weight_status", fill = "Weight_status", color = "white",
               palette = c("#00AFBB", "#E7B800", "#FC4E07"))
```

***The data reveals that 50% of the users are overweight while 35% are healthy and the other 15% are obese using their BMI status.***

### Determining relationship between sleep and activities
```{r}
activity_df <- daily_activity%>%
                  select(Id, Date, TotalSteps, VeryActiveMinutes, Calories)
colnames(activity_df)
colnames(sleepday)
```


### Merging two dataframes usig common columns
```{r}
activity_sleep <- merge(activity_df, sleepday, by = c("Id", "Date"))
head(activity_sleep)
```
### Adding a new column to cater for non sleeping time while in bed
```{r}
activity_sleep <- activity_sleep %>%
           mutate(NonSleepTimeInBed = TotalTimeInBed - TotalMinutesAsleep)
head(activity_sleep)
```

### Use correlation to check if there is any relationship
```{r}
corr_activity_sleep <- dplyr::select(activity_sleep,-c("Id","Date"))
corr <- round(cor(corr_activity_sleep),3)
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE)
```

***The correlation chart buttresses further the relationship between steps taken and calories expended by users on a day to day basics.***

***Also, people with active minutes has a strong correlation with calories burn.***

***There is no correlation between total minutes asleep and calories burn by users.***

### Checking for ActiveMinutes and Calories Burnt
```{r}
ggplot(data = activity_sleep) + geom_smooth(mapping = aes(x=VeryActiveMinutes, y=Calories)) +
  geom_point(mapping = aes(x=VeryActiveMinutes, y=Calories, color = VeryActiveMinutes)) +
  labs(title = "Calories Vs Very Active Minutes", x = "Very active minutes", y = "Calories")
```

***It reveals further that people with more active time burn more calories than than those with minimal time.***

### Checking for relationsship NonSleepTimeInBed and Calories
```{r}
ggplot(data = activity_sleep) + geom_smooth(mapping = aes(x=NonSleepTimeInBed, y=Calories)) +
  geom_point(mapping = aes(x=NonSleepTimeInBed, y=Calories), color = "green") +
  labs(title = "Calories Vs Non_sleep time in bed", x = "Non_sleep time in bed", y = "Calories")
```

### Merging the heartrate df to the activity_sleep data to know if any relationship exist amongst the data.
```{r}
heartrate <- heartrate %>%
            group_by(Id, Date) %>% 
                      summarise(daily_heartrate = mean(Value))
head(heartrate) 

activity_sleep_heartrate <- merge(activity_sleep, heartrate, by = c("Id", "Date"))
head(activity_sleep_heartrate)
```


### Checking the correlation
```{r}
corr_activity_sleep_heartrate <- dplyr::select(activity_sleep_heartrate,-c("Id","Date"))
corr <- round(cor(corr_activity_sleep_heartrate),3)
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE)
```

***The daily heartrate has no correlation with active minutes & total steps meanwhile shows a -ve correlation for total time in bed, minutes asleep & non sleep time in bed***

### Relationship between Total Minutes Asleep & daily heartrate
```{r}
ggplot(data = activity_sleep_heartrate) + geom_smooth(mapping = aes(x=TotalMinutesAsleep, y=daily_heartrate)) +
  geom_point(mapping = aes(x=TotalMinutesAsleep, y=daily_heartrate), color = "darkgreen") +
  labs(title = "Daily Heartrate Vs Total Minutes Asleep", x = "Total Minutes Asleep", y = "Daily Heartrate")
```

***The chart shows that despite the number of time spent asleep on a daily basis, has no impact on the heart beat rate.***

### Relationship between Total Minutes Asleep and Daily Heartrate
```{r}
ggplot(data = activity_sleep_heartrate) + geom_smooth(mapping = aes(x= NonSleepTimeInBed, y=daily_heartrate)) +
  geom_point(mapping = aes(x=NonSleepTimeInBed, y=daily_heartrate), color = "darkgreen") +
  labs(title = "Daily Heartrate Vs Non Sleep Time In Bed", x = "Non Sleep Time In Bed", y = "Daily Heartrate")
```

***No clear relationship existing between non sleeping time in bed and daily heartrate.***

## Key Findings

* The average calories expended on a daily basis  ranges between 2250 - 2400.

* 50% of customers using the product are overweight while 35% are healthy & the remaining 15% are obese using their BMI score.

* The more steps covered per day is directly proportional to calories expended which is reflective of users that are having more active hours.

* Subscribers with more active minutes burns more calories.

* Users recording more active minutes took more steps on a daily basis and expend more calories than those with more none active minutes.

* There is no positive correlation between daily heart rate & the following parameters; Total time in bed, Total minutes asleep & Non sleep time in bed.

## Recommendatons

* The data set available are limited to get a more comprehensive analysis and the need to gather more data set for a longer period of time is necessary.

* Customers can get a monthly report of their calories, health status using BMI or weight & other vitals as email to show accountability & progress they make. Perks like this can make Bella bet stand out.

* Create discount for users that introduce a new customer.

* Introduce products/activities that will target users that are overweight to burn more calories & create personalized services since most users are in this categories.

* Create more awareness across digital platforms & also work with health centers to have access to new mothers that will need the product.

* Quarterly reports should be made available to the public on media spaces for general awareness & benefits of the product.

